\keyword{C, formal semantics, compiler verification, relaxed-memory concurrency, separate
  compilation, integer-pointer cast}
\begin{abstract}
  To improve the performance of C programs, mainstream compilers perform aggressive optimizations
  that may change the behaviors of programs that use low-level features in unidiomatic ways.
  Unfortunately, despite many years of research and industrial efforts, it has proven very difficult
  to adequately balance the conflicting criteria for low-level features and compiler optimizations
  in the design of the C programming language.  On the one hand, C should support the common usage
  patterns of the low-level features in systems programming.  On the other hand, C should also
  support the sophisticated and yet effective optimizations performed by mainstream compilers.  None
  of the existing proposals for C semantics, however, sufficiently support low-level features and
  compiler optimizations at the same time.

  In this dissertation, we resolve the conflict between some of the low-level features crucially
  used in systems programming and major compiler optimizations.  Specifically, we develop the first
  formal semantics of relaxed-memory concurrency, separate compilation, and cast between integers
  and pointers that $(1)$ supports their common usage patterns and reasoning principles for
  programmers, and $(2)$ provably validates major compiler optimizations at the same time.  To
  establish confidence in our formal semantics, we have formalized most of our key results in the
  Coq theorem prover, which automatically and rigorously checks the validity of the results.

  % advance the formal semantics and compiler verification techniques for C to account for language
  % features and compiler use cases that were beyond the reach of the prior work and yet are crucial
  % in the real world.  By doing so, we enhance the understanding of the practice of C programming
  % language and its compilers.  Specifically, we make the following contributions.  First, we propose
  % a formal semantics of integer-pointer casts and relaxed-memory concurrency.  These features are
  % crucial in that they allow cross-platform low-level access to memory and are essential for
  % applications such as operating system kernels and language runtimes.  Second, we generalize
  % CompCert to support for separate compilation with low verification cost.  Separate compilation is
  % crucial in that it significantly reduces compilation time.  To establish confidence in our
  % contributions, we have formalized most of the key results in Coq.

  % To address these problems, programming language researchers have proposed to design formal
  % semantics and verify compilers for C.  Formal semantics is a mathematical definition of program
  % behaviors so that it can clearly decide whether optimization is right or wrong.  Verified
  % compilers are accompanied with machine-checked correctness proof so that they can be used for
  % safety-critical systems with confidence.  However, prior work on formal semantics and compiler
  % verification makes unrealistic simplifying assumptions on C semantics and compilers, skating over
  % the complexity of the real-world practice of systems programming.

\end{abstract}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% TeX-command-extra-options: "-shell-escape"
%%% End:
