\acknowledgement

Without great sacrifices of teachers, I would not have been able to finish my Ph.D.  Prof. Kwangkeun
Yi bore with my childish enthusiasm and tamed it into a dedication.  My advisor Prof. Chung-Kil Hur
taught me how to read, write, listen, speak, and think by doing together.  Dr. Derek Dreyer gave
constructive advice that I can follow at just the right times.  Now I realize everything they have
done to me requires a lot of patience.  I would like to say thank you with all my heart.

I would also like to thank Prof. Byung-Gon Chun and Prof. Jae W. Lee for reviewing and giving
feedback on this dissertation and my job talk as committee members.  I also thank William Mansky,
Dmitri Garbuzov, Steve Zdancewic, Viktor Vafeiadis, Yoonseung Kim, and Ori Lahav for working
together on the papers which this dissertation is based on, and thank Robbert Krebbers, Jim Apple,
Xavier Leroy, Doug Lea, Alan Jeffrey, Andreas Lochbihler, James Riely, Peter Sewell, and Joe
Tassarotti for very helpful feedback on the papers.

The papers on which this dissertation based are supported in part by the Engineering Research Center
of Excellence Program of MSIP/NRF (Grant NRF-2008-0062609), the ICT R\&D program of MSIP/IITP (Grant
R0132-15-1006), Samsung Research Funding Center of Samsung Electronics (Project Number
SRFC-IT1502-07), a Kwanjeong Educational Foundation Scholarship, and a Korea Foundation for Advanced
Studies Scholarship.


\paragraph*{}

\hyphenation{Jong-hwan}

I remember my friends.  Jaeyeol have always been my teacher, mentor, friend, and brother, and he
organized ``Kang Sa Mo'' for my wedding ceremony.  Minsuk and Jonghwan took care of me as if I'm
their younger brother.  Hyeong Kyun and I went through bad times together as comrades, and Joonhyuk,
Taehong, Hyunsung and I went through good times together as comrades.  I remember Wonha.  I also
remember all ROPAS and SF members, and especially Soonho, Wonchan, and Wontae for helping me in my
early career.


\paragraph*{}

None of this could have happened without my family.  Mom and Dad, you have always provided me with
all that I needed so far.  Now I am starting to realize what it meant for you only after having my
own child.  All I can say is just thank you and love you.  Grandma, you colored my childhood with
joy and warmth.  I miss you so much and I really hope you could attend my commencement ceremony.
Minjung, your positive thinking helped us bond together as a family, in good times and in bad.  You
are probably the greatest sister in the history.

And Eunjung, my lovely and brilliant wife, you sacrificed a lot of things for this dissertation and
my career.  Thank you for supporting and loving me.  This dissertation, as well as the rest of my
life, is dedicated to you.  Saeun, your smile raises me up even when I am totally exhausted.
Someday you will grow up, but I will be your shelter forever.  Eunjung and Saeun, I will always love
you no matter what.


% \jeehoon{Before submitting the thesis, check landscape pages.}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% TeX-command-extra-options: "-shell-escape"
%%% End:
